---
title: "Lesson 8 RMarkdown - Output formats"
---

```{r library}

```


<div class="question">

Exercise 8.1 

This exercise is adapted from [this one from r4ds here](https://r4ds.had.co.nz/r-markdown.html#exercises-71)

Copy and paste the contents of Lorem.Rmd <a href="https://github.com/DataScienceILC/tlsc-dsfb26v-20_workflows/raw/main/Rmd/lorem.Rmd" download> here </a> to a .Rmd file.

From this .Rmd, knit one new document for each of the three built-in formats: HTML, PDF and Word. How does the output differ? How does the input differ? (You may need to install LaTeX in order to build the PDF output — RStudio will prompt you if this is necessary.)
</div>

The output is different depending on the file format of course.
.html output results in .html files. Quarto also has the ability to render pages as pdf and LaTeX files, which requires the quarto options: `format: pdf` and `format: latex`. Another possibility is to use `format: pdf: keep-tex: true` to get pdf output and keep `.tex` for further editing.

Looks-wise html looks like a website document, pdf looks more similar too a book or paper, latex output results in a raw .tex file. All of these outputs also create their own _files folder containing graphics.

Some distribution of TeX is required to have R convert the output to TeX files and TinyTeX is recommended. 

PDF is a format which is a combination of text and graphics, TeX is more similar to markdown in that it is a language, which have differing syntax and capabilities. LaTeX is seen as the academic standard for its powerful precision over text/page layout and mathematical typesetting.
However, rmarkdown, combined with quarto and its packages, css and html allows for many options regarding many other aspects. Combined with its ease of use makes it a great tool for rendering websites, documents and allows for more interactivity over LaTeX' more static documents.

<div class="question">

Exercise 8.2

  A) What is a floating TOC?
  B) Use Lorem.Rmd again and add a TOC
  C) Give the TOC a set depth (toc_depth) and see what happens if you change it. (for instance `toc_depth: 2` can be set just as you set `toc_float: true` earlier.)
  D) Also try `code_folding: hide` and `number_sections: FALSE`.
  E) Google how to put two authors on one .Rmd, with affiliations. Put your and your labpartner's names as authors.
  F) Apply a theme you like, but not "united".
  G) try tabbed sections

</div>

```{r}
#| eval: false

# A-F

---
title: "lorem"
format: 
  html:
   toc: true
   toc-location: body
   toc-depth: 2
   code-fold: true
   number-sections: true
   theme: cosmo
author:
  - name: Alex Groot
    orcid: 1730201
    email: email@mail.com
    affiliations:
      - name: Hogeschool Utrecht
  - name: Author 2
    affiliations:
      - name: Hogeschool Utrecht
---
      
# Panel tabsets:
      
:::{.panel-tabset}
### header
    text
### header2
    text2
:::
      
```


<div class="question">
Exercise 8.3
 - Follow [these steps until "Beyond the basics".(click)](https://rmarkdown.rstudio.com/articles_docx.html)
 - Get your word-styles-reference-01.docx to work as a style reference.

</div>

```{r}
#| eval: false

---
title: "008_test_report"
format: 
  docx:
    reference-doc: "word-styles-reference-01.docx" 
---
  
```

The rendered document in "_site/008_test_report.docx" correctly uses the references-doc and is generated with different style, such as the title being "DejaVu Math TeX Gyre" in the new file using reference, while a new document created within LibreOffice utilizes the default title "Liberation Sans".

<div class="question">
Exercise 8.4

**rabbits!**

Change the markers in the example above to little icons of rabbits! See [here (click)](http://rstudio.github.io/leaflet/markers.html) how to do this.
</div>

```{r}
library(leaflet)
library(tidyverse)

rabbit_locations <- tibble( rabbitnr = c(1:5),
                            lat  = c(52.0852,52.0832,52.08247,52.0834,52.08309),
                            long = c(5.171109,5.171248,5.171248,5.176055,5.17268) )
# Download free icon from "https://fonts.google.com/icons"
# Edited to be coloured black
rabbit_icons <- icons(
  iconUrl = here::here("images/data_0080/cruelty_free_24dp_E8EAED_FILL0_wght400_GRAD0_opsz24.png"),
  iconWidth = 32,
  iconHeight = 32
)

leaflet(data = rabbit_locations) %>% addTiles() %>% 
  addMarkers(~long, ~lat, popup = ~as.character(rabbitnr), label = ~as.character(rabbitnr), 
             icon = rabbit_icons) # add icon
```

