---
title: Lesson 9 RMarkdown Parameters & Presentations
---
## ocean_floor.qmd

```{r}
#| eval: false

---
title: "Visualizing the ocean floor"
output: html_document
params:
  data: "florida"
---


library(marmap)
library(ggplot2)


The [marmap](https://cran.r-project.org/web/packages/marmap/index.html) package provides tools and data for visualizing the ocean floor. Here is an example contour plot of marmap's ``r params$data`` dataset.


print(params$data)
data(list = params$data)
autoplot.bathy(get(params$data))
```

The tutorial code can't `autoplot()` the data automatically as the input data is in `bathy` format.
Changing `autoplot()` to `autoplot.bathy()` enables the document to render the ocean floor map.

## ocean_floor_loop.R script
```{r}
#| eval: false
library(here)
library(rmarkdown)
library(tidyverse)
library(vcdExtra)

# load bathy dataset from marmap
param_list <-
  vcdExtra::datasets("marmap") %>% filter(class == "bathy") %>% select(Item)

for(data_input in param_list$Item){
  
  input_file <- here::here("ocean_floor.qmd")
  output_file <- paste0("009_param_", data_input, ".html")
  output_dir <- paste0("_site")
  
  rmarkdown::render(
    input = input_file,
    params = list(data = data_input),
    output_file = output_file,
    output_dir = output_dir
  )}
```

This script select the `bathy` data from the `marmap` dataset and loops over the parameter names and creates separate `.html` output for each in the `_site` folder. 

## Example output

```{r include = FALSE}
library(marmap)
library(ggplot2)
dummy_param <- "hawaii"
```

The [marmap](https://cran.r-project.org/web/packages/marmap/index.html) package provides tools and data for visualizing the ocean floor. Here is an example contour plot of marmap's ``r dummy_param`` dataset.

```{r echo = TRUE}
print(dummy_param)
data(list = dummy_param)
autoplot.bathy(get(dummy_param))
```